Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-02-18 agujournaltemplate.tex
Intro draft Day 2 of Focus

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low-upwelling conditions.
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

\begin{itemize}

    \item - Phytoplankton is important

    %“Time series of biological observations provide insights into the responses of marine ecosystems to cyclic processes, transient events, and, when maintained over many decades, changes in climate (Carstensen 2014, Henson et al. 2016). Such time series are needed for assessments on the dynamic state of the ocean.” ([Muller-Karger et al., 2019, p. 2](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=2&annotation=R4DSZPK6))

    - Phytoplankton is important
    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?

    - some general intro to DIVERSITY.. what is it? Why important?
    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al. 2013), the biodiversity response is much more complex. First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g. species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness). But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity. Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions. Second, for each of these facets, all metrics are effort- and scaledependent. Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management. But, the relevance of the metrics arises from analysing their temporal trends over a long time period. Third, no single variable captures even the most important aspects of community composition and change. Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al. 2023 (with Hillebrand)

    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
    \item "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change. Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
        -

    \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.
        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. 
        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}. 
        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m. The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.

@@

    \item "The CARIACO Time Series ...." - talk about the Time Series, what they did, why

    \item "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature

@@

    \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
    - what has been done (Phytoplankton - Statistics)?
    - where are the gaps?

    So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
    Or nutrients, but if it is a climate indices, then the indices capture some variablity that is not related to wind or nutrients....

    \item "In this paper, we..."

        %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.

    \end{itemize}

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
\caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance accross predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:GF}
\end{figure}

@@

@@

@@

>>>>

\begin{keypoints}
\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017. Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index. 
% \index Previous studies proposed a shift in the community, but with the full dataset we see a shift and a return. The community shows a departure form cluster one around 2004, but a return to that community in 2014. 
\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem. The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind. 
\end{keypoints}

@@

    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}. These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014,henson_observing_2016}. 
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin off the coast of Venezuela between 1995 and 2017 \cite{muller-karger_scientific_2019}. The Cariaco basin is one of the most productive ecosystems in the Caribbean Sea \cite{bringue_physico-chemical_2018}, with seasonal upwelling driven by trade winds during December through May and a secondary upwelling observed in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is anoxic at depths below \qty{\sim 275}{m}, due to microbial respiration of sinking organic matter and restricted bottom water circulation, leaving a rich sedimentary record for the study of past climate \cite{hughen1996nature}.

    % In recent years, significant changes have been observed in phytoplankton populations within the Cariaco Basin. These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns. Understanding the drivers and impacts of these changes in the phytoplankton community is crucial to understand the broader implications for marine food webs and carbon cycling in this and similar ecosystems.

    "The Cariaco Basin, located ...." - general hydrography/depth, etc.
        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. 
        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}. 
        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m. The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.

@@

    "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature

@@

    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?

    - some general intro to DIVERSITY.. what is it? Why important?

    % "Phytoplankton is the major primary producer group in marine ecosystems and at the same time a highly suitable indicator for ecosystem change. Due to its high diversity and short generation time, phytoplankton responds quickly to changes in the environment (Reynolds 1998). Especially changes in temperature and nutrient supply are known to change phytoplankton composition and biodiversity, with higher temperatures and lower nutrients favouring small species (Litchman et al. 2007; Winder and Sommer 2012), whereas Si:N and N:P supply ratios change the relative importance of diatoms and other groups such as dino(flagellates) and cyanobacteria (Sommer et al. 2004; Makareviciute-Fichtner et al. 2020)." from Di Cavalho et al. 2023

    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al. 2013), the biodiversity response is much more complex. First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g. species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness). But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity. Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions. Second, for each of these facets, all metrics are effort- and scaledependent. Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management. But, the relevance of the metrics arises from analysing their temporal trends over a long time period. Third, no single variable captures even the most important aspects of community composition and change. Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al. 2023 (with Hillebrand)

    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
    "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change. Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
        -

        "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
    - what has been done (Phytoplankton - Statistics)?
    - where are the gaps?
    - less studied aspects: Diversity + Phytoplankton Microscopy

    So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
    Or nutrients, but if it is a climate indices, then the indices capture some variability that is not related to wind or nutrients....

    "In this paper, we..."

        %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.
        %What are the questions I ask myself?? 1.

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
\caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
\label{fig:GF}
\end{figure}

@@

- The 20 year record of CARIACO time series is a great opportunity to better understand the connection between pelagic proccesses, climate and sedimentation.

- Our main results show that...

- We are not able to fully resolve....

- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
%It is well known that the personal component has an influence on how well species are identified. Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al. 2012; Nohe et al. 2018). A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved. Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1). In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al. 2023). % from Di Cavalho et al. 2023

- The decrease in observed phytoplankton cells between the two clusters indicates...
% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups

- What could have driven these shifts?
% talk about link to bottom up drivers, but also mention top down aspects... then go deep on Gradient Forest analysis, what that could mean. Relate this to other papers about community changes, climate effects and gradient forest applications.

@@

% Pinckney et al. 2015:
- major thing is that previous paper only looked at microscopy to 55m (Mutshinda and Pinckney), but now we look at full 100 m!
%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014). Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014). This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012). Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013). During this period, a collapse in the sardine industry also occurred. The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012). These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
% Consistent with the results of Taylor et al. (2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m. However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3). We argue that the reduction in cell numbers is likely only true of larger cells. Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000). Furthermore, Mousing et al. (2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability. Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e. below 56 m). In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years. In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014). Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.

@@

- Our analysis shows that...

- We also show that...

- But we cannot confirm or test. so we would recommend X for the future.
